Building a GNU/Hurd Cluster

### 2018-01-17 papers/cua.tex
cua lecture: version of the slides for the presentation

\usepackage{tikz}
\usetikzlibrary{positioning, fit, backgrounds, arrows, patterns}

@@

\begin{frame}
\frametitle{Techniques to exploit parallelism}
\begin{itemize}
\item Instruction level parallelism
\begin{itemize}
\item Superscalar vector operations
\item Speculative execution
\item Co-processors
\end{itemize}
\item Thread level parallelism
\begin{itemize}
\item Shared memory
\item Caching
\item Synchronization
\end{itemize}
\item Node level parallelism
\begin{itemize}
\item No shared memory
\item High speed LAN
\item Clusters
\end{itemize}
\end{itemize}
\end{frame}

@@

@@

\begin{frame}
\frametitle{Cluster Computing - History}
\begin{itemize}
\item Beowulf
\item Hadoop
\item Cluster Linux
\begin{itemize}
\item Kerrighed - Latest version is 3.0.0 which was released on 14 June 2010 and based on Linux kernel 2.6.30.
\item OpenSSI - last development release 2010; last stable release 2007
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Key Features
\begin{itemize}
\item Cluster-wide login
\item Single Process Space
\item Process Migration
\item Single Root Filesystem
\item Redundant Filesystem
\item Single I/O Space - block devices
\item Single IPC Space
standard Linux inter-process communication mechanisms, shared memory, semaphores, SYSV message queues, pipes and Unix domain sockets.
\item Cluster-wide IP Addresses
\end{itemize}
\item Advantages
\begin{itemize}
\item familiar programming and management tools
\item ease of leveraging additional computing power
\end{itemize}
\item Disadvantages
\begin{itemize}
\item less fault isolation
\item node failure more difficult
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Single System Image - History}
\begin{itemize}
\item VMS
\item SCALEMP
\item Cluster Linux
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{GNU Hurd}
\begin{itemize}
\item History
\item Microkernels in general
\item CMU Mach
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{CMU Mach}
\begin{itemize}
\item Tasks / Threads
\item Interprocess Communication
\item Memory Management
\end{itemize}
\end{frame}

@@

% \task - takes four arguments
% 1. name of task (used to refer to it from tikz; maybe printed as a label as well)
% 2. location of task box's center point
% 3. "r" or "l", indicating if a set of tick marks representing
%    ports should drawn on the left or right side of the task box
%    "u" to label the task "Task" instead of labeling it with its name
%    "b" to make a "big" node (I wish I knew how to size the tick marks automatically)
% 4. contents of the task box - other nodes
%
% The tick marks will be named "#1 Port #" for future reference

@@

\foreach \i in {-7,...,7}
{
\IfSubStr{#3}{l}{
  \draw[xshift=-5em, ultra thick] (0,0.2 * \i)  coordinate (#1 Port \i) -- (0.2,0.2 * \i);
}{
  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i);
}
\IfSubStr{#3}{r}{
  \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
}{
  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i);
}
}

\newcommand\bigtask[4]{
\matrix (#1) [task, inner xsep=0] at (#2) { \node {\IfSubStr{#3}{u}{Task}{#1}}; \\ #4

\foreach \i in {-20,...,7}
{
\IfSubStr{#3}{l}{
  \draw[xshift=-5em, ultra thick] (0,0.2 * \i)  coordinate (#1 Port \i) -- (0.2,0.2 * \i);
}{
  \draw[xshift=-5em, ultra thick, color=blue!20] (0,0.2 * \i) -- (0.2,0.2 * \i);
}
\IfSubStr{#3}{r}{
  \draw[xshift=5em, ultra thick] (-0.2,0.2 * \i) -- (0,0.2 * \i) coordinate (#1 Port \i);
}{
  \draw[xshift=5em, ultra thick, color=blue!20] (-0.2,0.2 * \i) -- (0,0.2 * \i);
}
}

@@

\bigtask{Task 1}{0,0}{r}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright
\only<2->{
\node (sendright) [draw, fill=orange] {Send 41};
\node (message1) [fit=(sendright), message] {Message 1};
\node (sendright) [draw, fill=orange] {Send 41};
}

@@

\bigtask{Task 2}{15 em,0}{l}{
\only<3->{
\node (sendright2) [draw, fill=orange] {Send 9};
\node (message2) [fit=(sendright2), message] {Message 1};
\node (sendright2) [draw, fill=orange] {Send 9};
}

@@

\bigtask{Task 1}{0,0}{r}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright
\only<2->{
\node (sendright) [draw, fill=green] {Recv 41};
\node (message1) [fit=(sendright), message] {Message 1};
\node (sendright) [draw, fill=green] {Recv 41};
}

@@

\bigtask{Task 2}{15 em,0}{l}{
\only<3->{
\node (sendright2) [draw, fill=green] {Recv 9};
\node (message2) [fit=(sendright2), message] {Message 1};
\node (sendright2) [draw, fill=green] {Recv 9};
}

@@

\bigtask{Task 1}{0,0}{r}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright
\node (message1) [message] {Message};
\only<1>{
\node (memory1) [below=1em, fill=orange!40, minimum size=2cm, dashed, thick, draw] at (message1.south) {Memory};
\draw [thick, color=red, dotted, -] (message1) -- (memory1);
}
}

\bigtask{Task 2}{15 em,0}{l}{
\only<2->{
\node (message2) [message] {Message};
\node (memory2) [below=1em, fill=orange!40, minimum size=2cm, dashed, thick, draw] at (message2.south) {Memory};
\draw [thick, color=red, dotted, -] (message2) -- (memory2);
}
}

@@

\begin{frame}
\frametitle{Hurd Filesystem}
\end{frame}

@@

\begin{frame}
\frametitle{Example: Basic File I/O}
\end{frame}

\draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 8 em);
\draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 8 em);

\begin{frame}
\frametitle{Example: Shared Libraries}
\end{frame}

@@

\begin{frame}
\frametitle{The {\tt netmsg} Server}
\end{frame}

@@

\begin{frame}
\frametitle{What kind of bugs have I found?}
\begin{itemize}
\item failure to update pagemap entry
\item FLUSHING flag
\item DEAD NAME notifications
\item condition variable
\item weak refs
\end{itemize}
\end{frame}

>>>>

\usepackage{tikz}
\usetikzlibrary{positioning, fit, backgrounds, arrows, patterns, trees, shapes}

@@

\begin{frame}
\frametitle{Techniques to exploit parallelism (partial list)}
\begin{itemize}
\item Instruction level parallelism
\begin{itemize}
\item Superscalar vector operations
\end{itemize}
\item Thread level parallelism
\begin{itemize}
\item Multiple cores with shared memory
\end{itemize}
\item Co-processors
\begin{itemize}
\item More cores; simpler cores
\end{itemize}
\item Node level parallelism
\begin{itemize}
\item Multiple cores without shared memory
\item High speed LAN
\item Clusters
\end{itemize}
\end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Adding 512 64-bit integers -- Multi-threaded Code}
\begin{semiverbatim}
\tiny
#include <thread>

int a[512], b[512], c[512];

int operation(int * ap, int * bp, int * cp, int count)
\{
  for (int i = 0; i < count; i++) \{
    cp[i] = ap[i] + bp[i];
  \}
\}

int main () \{
  const int num_threads = 3;
  int start_index[num_threads+1];
  std::thread thread[num_threads];

  start_index[0] = 0;
  for (int i=1; i < num_threads; i++) \{
    start_index[i] = i*(512/num_threads) & ~15;
  \}
  start_index[num_threads] = 512;

  for (int i=0; i < num_threads; i++) \{
    thread[i] = std::thread(operation,
                            &a[start_index[i]], &b[start_index[i]], &c[start_index[i]],
                            start_index[i+1] - start_index[i]);
  \}

  for (int i=0; i < num_threads; i++) \{
    thread[i].join();
  \}
\}
\end{semiverbatim}
\end{frame}

@@

\begin{frame}[fragile]
\frametitle{Adding 512 64-bit integers -- OpenCL kernel}
\begin{semiverbatim}
\tiny
void kernel simple_add(global const int* A, global const int* B, global int* C,
                       global const int* N) \{
    int ID, Nthreads, n, ratio, start, stop;

    ID = get_global_id(0);
    Nthreads = get_global_size(0);
    n = N[0];

    ratio = (n / Nthreads);  // number of elements for each thread
    start = ratio * ID;
    stop  = ratio * (ID + 1);

    for (int i=start; i<stop; i++)
        C[i] = A[i] + B[i];
\}
\end{semiverbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{Adding 512 64-bit integers -- OpenCL program}
\begin{semiverbatim}
\tiny
const int n = 512;
int A[n], B[n], C[n];

cl::Program program(context, sources);
program.build({default_device});

int N[1] = {n};

// create buffers on device (allocate space on GPU)
cl::Buffer buffer_A(context, CL_MEM_READ_WRITE, sizeof(int) * n);
cl::Buffer buffer_B(context, CL_MEM_READ_WRITE, sizeof(int) * n);
cl::Buffer buffer_C(context, CL_MEM_READ_WRITE, sizeof(int) * n);
cl::Buffer buffer_N(context, CL_MEM_READ_ONLY,  sizeof(int));

// create a queue (a queue of commands that the GPU will execute)
cl::CommandQueue queue(context, default_device);

// push write commands to queue
queue.enqueueWriteBuffer(buffer_A, CL_TRUE, 0, sizeof(int)*n, A);
queue.enqueueWriteBuffer(buffer_B, CL_TRUE, 0, sizeof(int)*n, B);
queue.enqueueWriteBuffer(buffer_N, CL_TRUE, 0, sizeof(int),   N);

cl::Kernel simple_add(program, "simple_add");
simple_add.setArg(0, buffer_A);
simple_add.setArg(1, buffer_B);
simple_add.setArg(2, buffer_C);
queue.enqueueNDRangeKernel(simple_add, cl::NullRange, cl::NDRange(20), cl::NullRange);

// read result from GPU to here
queue.enqueueReadBuffer(buffer_C, CL_TRUE, 0, sizeof(int)*n, C);

\begin{frame}
\frametitle{Techniques to exploit parallelism (partial list)}
\begin{itemize}
\item Instruction level parallelism
\begin{itemize}
\item Superscalar vector operations
\item \alert{Short {\tt for} loops}
\end{itemize}
\item Thread level parallelism
\begin{itemize}
\item Multiple cores with shared memory
\item \alert{Synchronization}
\end{itemize}
\item Co-processors
\begin{itemize}
\item More cores; simpler cores
\item \alert{OpenCL}
\end{itemize}
\item Node level parallelism
\begin{itemize}
\item Multiple cores without shared memory
\item High speed LAN
\item Clusters
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Cluster Computing - Partial History}
\begin{itemize}
\item VMS
\item Beowulf
\begin{itemize}
\item Pioneered in 1994 at NASA
\item Message Passing Interface (MPI)
\item Parallel Virtual Machine (PVM)
\end{itemize}
\item Hadoop
\item Single System Image
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item Key Features
\begin{itemize}
\item Cluster-wide login
\item Single Process Space
\item Process Migration
\item Single Root Filesystem
\item Redundant Filesystem
\item Single I/O Space - block devices
\item Single IPC Space
standard Linux inter-process communication mechanisms, shared memory, semaphores, SYSV message queues, pipes and Unix domain sockets.
\item Cluster-wide IP Addresses
\end{itemize}
\item Advantages
\begin{itemize}
\item familiar programming and management tools
\item ease of leveraging additional computing power
\end{itemize}
\item Disadvantages
\begin{itemize}
\item less fault isolation
\item node failure more difficult to handle
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{POSIX System System Image -- Partial History}
\begin{itemize}
\item Single System Image Linux
\begin{itemize}
\item MOSIX
\begin{itemize}
\item Developed at Hebrew University at Jerusalem since 1977
\item Closed source; free download
\item Latest version works with Linux 4.12.13; released in 2017
\end{itemize}
\item OpenMosix
\begin{itemize}
\item Forked from last free version of MOSIX in 2001
\item Last development release 2006 (based on Linux 2.6.15)
\item Last stable release 2004 (based on Linux 2.4.26; 38,794 line diff)
\item project terminated in 2008
\end{itemize}
\item Kerrighed
\begin{itemize}
\item {\it Institut national de recherche en informatique et en automatique}
\item Latest version (3.0.0) released on 14 June 2010
\item Based on Linux 2.6.30
\end{itemize}
\item OpenSSI
\begin{itemize}
\item Funded by SCO; Compaq
\item Last development release 2010 (based on Linux 2.6.12)
\item Last stable release 2007 (based on Linux 2.6.11)
\end{itemize}
\end{itemize}
\item ScaleMP
% \item CMU Mach / GNU Hurd
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{CMU Mach / GNU Hurd}
\begin{itemize}
\item{CMU Mach}
\begin{itemize}
\item Developed at Carnegie Mellon University from 1985 to 1994
\item Pioneered the Microkernel concept
\item Kernel does Tasks / Threads / IPC / Memory Management
\item Everything else in user space (device drivers; filesystem; networking)
\item Designed for multi-node clusters
\end{itemize}
\item{GNU Hurd}
\begin{itemize}
\item Under development since 1990
\item Runs on top of Mach and implements a full POSIX system
\item Averages about two releases per year
\item Available as part of Debian
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{How many Debian packages build on Hurd?}
\includegraphics[width=\textwidth]{graph-big.png}
\end{frame}

@@

% \task - takes four arguments
% 1. name of task (used to refer to it from tikz; maybe printed as a label as well)
% 2. location of task box's center point
% 3. "r" or "l", indicating if a set of tick marks representing
%    ports should drawn on the left or right side of the task box
%    "u" to label the task "Task" instead of labeling it with its name
%    "b" to make a "big" node (I wish I knew how to size the tick marks automatically)
%    "s" to make a "small" node
% 4. contents of the task box - other nodes
%
% The tick marks will be named "#1 Port #" for future reference
%
% It's implemented as a tikz matrix with two cells - the task name in the top cell
% and everything else in the bottom cell

@@

\IfSubStr{#3}{b}{\edef\starti{-20}}{\IfSubStr{#3}{s}{\edef\starti{-1}}{\edef\starti{-7}}}
\IfSubStr{#3}{s}{\edef\endi{1}}{\edef\endi{7}}

\foreach \i in {\starti,...,\endi}
{
  \IfSubStr{#3}{l}{\edef\leftcolor{black}\edef\leftcoord{coordinate (#1 Port \i)}}
                  {\edef\leftcolor{blue!20}\edef\leftcoord{}}
  \IfSubStr{#3}{r}{\edef\rightcolor{black}\edef\rightcoord{coordinate (#1 Port \i)}}
                  {\edef\rightcolor{blue!20}\edef\rightcoord{}}
  \draw[xshift=-5em, ultra thick, color=\leftcolor] (0,0.2 * \i) \leftcoord -- (0.2,0.2 * \i);
  \draw[xshift=5em, ultra thick, color=\rightcolor] (-0.2,0.2 * \i) -- (0,0.2 * \i) \rightcoord;
}

@@

\task{Task 1}{0,0}{br}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright
\only<2->{
\node (sendright) [draw, fill=orange] {Send 41};
\node (message1) [fit=(sendright), message] {Message 1};
\node (sendright) [draw, fill=orange] {Send 41};
}

@@

\task{Task 2}{15 em,0}{bl}{
\only<3->{
\node (sendright2) [draw, fill=orange] {Send 9};
\node (message2) [fit=(sendright2), message] {Message 1};
\node (sendright2) [draw, fill=orange] {Send 9};
}

@@

\task{Task 1}{0,0}{br}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright
\only<2->{
\node (sendright) [draw, fill=green] {Recv 41};
\node (message1) [fit=(sendright), message] {Message 1};
\node (sendright) [draw, fill=green] {Recv 41};
}

@@

\task{Task 2}{15 em,0}{bl}{
\only<3->{
\node (sendright2) [draw, fill=green] {Recv 9};
\node (message2) [fit=(sendright2), message] {Message 1};
\node (sendright2) [draw, fill=green] {Recv 9};
}

@@

\task{Task 1}{0,0}{br}{
% print two sendright nodes, one to fill the message1 box to the correct size,
% and another one to actually draw it, since message1 will overwrite the
% first sendright
\node (message1) [message] {Message};
\only<1>{
\node (memory1) [below=1em, fill=orange!40, minimum size=2cm, dashed, thick, draw] at (message1.south) {Memory};
\draw [thick, color=red, dotted, -] (message1) -- (memory1);
}
}

\task{Task 2}{15 em,0}{bl}{
\only<2->{
\node (message2) [message] {Message};
\node (memory2) [below=1em, fill=orange!40, minimum size=2cm, dashed, thick, draw] at (message2.south) {Memory};
\draw [thick, color=red, dotted, -] (message2) -- (memory2);
}
}

@@

\begin{frame}
\frametitle{Hurd Filesystem}
\tikzstyle{dirent}=[draw=black,thick,anchor=west]
\tikzstyle{selected}=[draw=red,fill=red!30]
\tikzstyle{optional}=[dashed,fill=gray!50]
\tikzstyle{link}=[red,dotted,->,thick]
\tikzstyle{grow2}=[grow via three points={one child at (1.5,0) and two children at (1.5,0) and (1.5,-0.7)},
  edge from parent path={(\tikzparentnode.east) -- +(.5,0) |- (\tikzchildnode.west)}]
\begin{center}
\begin{tikzpicture}
  [grow via three points={one child at (0.5,-0.7) and
  two children at (0.5,-0.7) and (0.5,-1.4)},
  edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
  \node[dirent] (root) {/}
  child { node[dirent] {bin}
    child [grow2] { node[dirent] (bash) {bash}}
    child [grow2] { node[dirent] (date) {date}}
    child [grow2] { node[dirent] (grep) {grep}}
    child [grow2] { node[dirent] (ls) {ls}}
  }
  child { node[dirent] {etc}}
  child { node[dirent] {lib}}
  child { node[dirent] {tmp}}
  child { node[dirent] (hurd) {hurd}}
  child { node[dirent] (home) {home}}
  child { node[dirent] (http) {http:}}
  child { node[dirent] {servers}
    child { node[dirent] {socket}
      child { node[dirent] (inet) {inet}}
    }
  };

\task{ext2 File Server}{20 em,-5 em}{l}{}
\task{httpfs}{20 em, -12 em}{ls}{}
\task{pfinet}{20 em, -17 em}{ls}{}

\draw [link] (root) -- +(5,0) |- (ext2 File Server Port 7);
\draw [link] (bash) -- +(1,0) |- (ext2 File Server Port 5);
\draw [link] (date) -- +(1,0) |- (ext2 File Server Port 3);
\draw [link] (grep) -- +(1.5,0) |- (ext2 File Server Port 1);
\draw [link] (ls) -- +(2,0) |- (ext2 File Server Port -1);
\draw [link] (hurd) -- +(2,0) -- (ext2 File Server Port -3);
\draw [link] (home) -- +(3,0) -- (ext2 File Server Port -5);

\draw [link] (http) -- +(2,0) |- (httpfs Port 0);
\draw [link] (inet) -- (pfinet Port -1);

@@

%\begin{frame}
%\frametitle{Example: Basic File I/O}
%\end{frame}

@@

\only<1>{
  \draw [color=red, <-, thick] (-2 em,-3 em) -- node [midway, left] {m\_o\_init} (-2 em,- 8 em);
  \draw [color=red, ->, thick] (-1.5 em,-3 em) -- node [midway, right] {m\_o\_ready} (-1.5 em,- 8 em);
}
\only<2>{
  \draw [color=red, <-, thick] (0.5 em,-3 em) -- node [midway, left] {m\_o\_data\_request} (0.5 em,- 8 em);
  \draw [color=red, ->, thick] (1 em,-3 em) -- node [midway, right] {m\_o\_data\_supply} (1 em,- 8 em);
}

\begin{frame}
\frametitle{Mach Memory Management: Shared Memory}
%\begin{center}
\begin{tikzpicture}[transform canvas={xshift=9.5em, yshift=2em}]

\task{Task 1}{-5em,0}{}{
\node (memory1) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
}
\task{Task 2}{6 em,0}{}{
\node (memory2) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
}
\task{Task 3}{17 em,0}{}{
\node (memory3) [fill=orange!40, minimum size=1.5cm, dashed, thick, draw] {};
}

\node (kernel) [kernel] at (6 em, -10 em) {};
\node (kmemory) [fill=orange!40, minimum size=1.5cm, thick, draw] at (6 em, -10 em) {};
\node [below] at (kernel.north) {Kernel};

\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory1);
\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory2);
\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory3);

\end{tikzpicture}
%\end{center}
\end{frame}

\begin{frame}
\frametitle{Mach Memory Management: Distributed Shared Memory}

\tikzstyle{emptymemory} = [fill=white, minimum size=1.5cm, dashed, thick, draw]
\tikzstyle{readmemory} = [fill=orange!40, minimum size=1.5cm, thick, draw]
\tikzstyle{writememory} = [fill=green!40, minimum size=1.5cm, thick, draw]

\begin{tikzpicture}[transform canvas={xshift=12em, yshift=7em, scale=0.4}]

\task{Task 1}{-5em,0}{u}{
\only<1-2,9->{\node (memory1) [emptymemory] {\only<2>{\Huge R}};}
\only<3-8> {\node (memory1) [readmemory, dashed] {};}
}
\task{Task 2}{6 em,0}{u}{
\only<1-3,9->{\node (memory2) [emptymemory] {};}
\only<4-8>{\node (memory2) [readmemory, dashed] {};}
}
\task{Task 3}{17 em,0}{u}{
\only<1-3,9-13>{\node (memory3) [emptymemory] {\only<11->{\Huge W}};}
\only<4-8>{\node (memory3) [readmemory, dashed] {};}
\only<14->{\node (memory3) [writememory, dashed] {};}
}

\node (kernel) [kernel] at (6 em, -10 em) {};
\only<1-2,9-13>{\node (kmemory) [fill=white, minimum size=1.5cm, dashed, thick, draw] at (6 em, -10 em) {};}
\only<3-8>{\node (kmemory) [fill=orange!40, minimum size=1.5cm, thick, draw] at (6 em, -10 em) {};}
\only<14>{\node (kmemory) [writememory] at (6 em, -10 em) {};}
\node [below] at (kernel.north) {Kernel};

\only<3-8>{\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory1);}
\only<4-8>{\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory2);}
\only<4-8>{\draw [orange, ultra thick, dotted, ->] (kmemory) -- (memory3);}
\only<14->{\draw [green, ultra thick, dotted, ->] (kmemory) -- (memory3);}

\node (server) [kernel, color=cyan!20, minimum width=20em] at (6 em, -20 em) {};
\only<1-9>{\node (smemory) [readmemory] at (server) {};}
\only<10-12,14>{\node (smemory) [emptymemory] at (server) {};}
\only<13>{\node (smemory) [writememory] at (server) {};}
\node [below] at (server.north) {Memory Object Server};

\only<2>{\draw [red, ->] (kernel) -- node [midway, left] {\huge m\_o\_data\_request} (server);}
\only<3>{\draw [red, <-] (kernel) -- node [midway, right] {\huge m\_o\_data\_supply} (server);}
\only<8>{\draw [red, <-] (kernel) -- node [midway, right] {\huge m\_o\_lock\_request (flush)} (server);}
\only<9>{\draw [red, ->] (kernel) -- node [midway, right] {\huge m\_o\_lock\_completed} (server);}
\only<11>{\draw [red, ->] (kernel) -- node [midway, left] {\huge m\_o\_data\_request} (server);}
\only<14>{\draw [red, <-] (kernel) -- node [midway, right] {\huge m\_o\_data\_supply} (server);}

\node (kernel2) [kernel] at (6 em, -30 em) {};
\only<1-5,13->{\node (kmemory2) [emptymemory] at (kernel2) {};}
\only<6-9>{\node (kmemory2) [readmemory] at (kernel2) {};}
\only<10-12>{\node (kmemory2) [writememory] at (kernel2) {};}
\node [below] at (kernel2.north) {Kernel};

\task{Task 4}{-5em,-40 em}{u}{
\node (memory4) [emptymemory] {};
}
\task{Task 5}{6 em,-40 em}{u}{
\only<1-5,13->{\node (memory5) [emptymemory] {\only<5>{\Huge R}};}
\only<6-12>{\node (memory5) [readmemory, dashed] {};}
}
\task{Task 6}{17 em,-40 em}{u}{
\only<1-9,13->{\node (memory6) [emptymemory] {\only<7-9>{\Huge W}};}
\only<10-12>{\node (memory6) [writememory, dashed] {};}
}

\only<5>{\draw [red, ->] (kernel2) -- node [midway, left] {\huge m\_o\_data\_request} (server);}
\only<6>{\draw [red, <-] (kernel2) -- node [midway, right] {\huge m\_o\_data\_supply} (server);}
\only<7>{\draw [red, ->] (kernel2) -- node [midway, left] {\huge m\_o\_data\_unlock} (server);}
\only<10>{\draw [red, <-] (kernel2) -- node [midway, right] {\huge m\_o\_lock\_request (+write)} (server);}
\only<12>{\draw [red, <-] (kernel2) -- node [midway, right] {\huge m\_o\_lock\_request (flush)} (server);}
\only<13>{\draw [red, ->] (kernel2) -- node [midway, left] {\huge m\_o\_data\_return} (server);}

\only<6-12>{\draw [orange, ultra thick, dotted, ->] (kmemory2) -- (memory5);}
\only<10-12>{\draw [green, ultra thick, dotted, ->] (kmemory2) -- (memory6);}

\end{tikzpicture}
\end{frame}

\begin{frame}
\frametitle{The {\tt netmsg} Server}
\begin{tikzpicture}[transform canvas={xshift=12em, yshift=6em, scale=0.4}]

\task{Task 1}{-5em,0}{u}{}
\task{Task 2}{6 em,0}{u}{}
\task{Task 3}{17 em,0}{u}{}

\node (kernel) [kernel] at (6 em, -10 em) {};
\node [above of=kernel] {\Huge netmsg};

\node [cloud, draw,cloud puffs=10,cloud puff arc=120, aspect=2, inner ysep=1em, minimum width=20em, minimum height=10em]
  at (6 em, -20 em) {};

\node (kernel2) [kernel] at (6 em, -30 em) {};
\node [above of=kernel2] {\Huge netmsg};

\task{Task 4}{-5em,-40 em}{u}{}
\task{Task 5}{6 em,-40 em}{u}{}
\task{Task 6}{17 em,-40 em}{u}{}

\draw [red] (Task 1.south) -- (kernel.center) -- (kernel2.center) -- (Task 4.north);
\draw [red] (Task 2.south) -- (kernel.center) -- (kernel2.center) -- (Task 5.north);
\draw [red] (Task 3.south) -- (kernel.center) -- (kernel2.center) -- (Task 6.north);

\end{tikzpicture}
\end{frame}

\begin{frame}
\frametitle{The {\tt netmsg} Server}
\begin{itemize}
\item All Mach messages serialized over a single TCP connection
\item Transmit Mach messages (almost) unchanged
\item Memory regions are appended at the end of the message
\item Local port numbers are transmitted unchanged
\item Remote port numbers have all their bits inverted
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The {\tt netmsg} Server}
\begin{center}
\begin{tikzpicture}[transform canvas={xshift=0em, yshift=-3em, scale=0.4}]

\fill[color=yellow!20] (-50 em, -10 em) rectangle (-8 em, 25 em);
\node at (-21 em, 20em) {\Huge Node 1};
\fill[color=yellow!20] (50 em, -10 em) rectangle (8 em, 25 em);
\node at (21 em, 20em) {\Huge Node 2};

\filldraw (-30 em, 15em) node[rectangle, name=queue 1, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
% \task{Task 0}{-30 em,15 em}{ur}{}

\task{Task 1}{-30 em,0}{r}{
  \only<1-2>{
    \node (sendright) [draw, fill=orange, inner xsep=5] {Send};
    \node (message1) [fit=(sendright), message] {Message};
    \only<1>{\node (sendright) [draw, fill=orange, inner xsep=5] {Send};}
  }
}
\task{netmsg 1}{-15 em,0}{l}{
  \only<2-3>{
    \node (sendright2) [draw, fill=orange, inner xsep=5] {Send};
    \node (message2) [fit=(sendright2), message] {Message};
  }
  \only<2->{\node (sendright2) [draw, fill=orange, inner xsep=5] {Send};}
}

\only<1>{\draw[red,dotted,->,ultra thick] (sendright) -- (queue 1);}
\only<2>{\draw[ultra thick,->] (message1) -- (message2);}
\only<2->{
  \draw[red,dotted,->,ultra thick] (sendright2) -- (queue 1);
}

\node [cloud, draw,cloud puffs=10,cloud puff arc=120, aspect=2, inner ysep=1em, minimum width=15em, minimum height=8em]
  at (0 em, 0 em) {};

\only<4->{
  \filldraw (15 em, 15em) node[rectangle, name=queue 2, minimum width=1.8cm, minimum height=1cm, draw, pattern=vertical lines] {};
}

\task{netmsg 2}{15 em,0}{r}{
  \only<3-5>{
    \node (sendright3) [draw, fill=orange, inner xsep=5] {Send};
    \node (message3) [fit=(sendright3), message] {Message};
    \only<4>{\node (sendright3) [draw, fill=orange, inner xsep=5] {Send};}
  }
  \only<4->{\node (rcvright3) [draw, fill=green, inner xsep=5] at (0,1.25) {Rcv};}
}
\task{Task 2}{30 em,0}{l}{
  \only<5>{
    \node (sendright4) [draw, fill=orange, inner xsep=5] {Send};
    \node (message4) [fit=(sendright4), message] {Message};
  }
  \only<5-6>{\node (sendright4) [draw, fill=orange, inner xsep=5] {Send};}
}

\only<1>{\draw[ultra thick, dotted, ->] (Task 1) -- (Task 2);}

\only<3>{\draw[ultra thick,->] (message2) -- (message3);}

\only<4>{\draw[red,dotted,->,ultra thick] (sendright3.west) to [bend left=20] (queue 2);}
\only<4->{\draw[green,dotted,->,ultra thick] (queue 2) -- (rcvright3);}

\only<5->{

\draw[red,dotted,->,ultra thick] (sendright4) -- (queue 2);
}

\only<5>{\draw[ultra thick,->] (message3) -- (message4);}

\begin{frame}[fragile]
\frametitle{{\tt netmsg}: Translation of a send right for transmission}
\begin{semiverbatim}
\tiny
if ((ports[i] == MACH_PORT_NULL) || (ports[i] == MACH_PORT_DEAD))
  \{
    continue;
  \}
else if (remote_ports_by_local.count(ports[i]) == 1)
  \{
    /* We're transmitting a send right that we earlier received over the network.  Convert to the
     * remote's name space, and indicate to the remote, by bit flipping the port number, that we're
     * sending it a port number in its own name space.  Also, destroy the send right we received
     * with the message, which is a send right to ourself!  If it's the port's last send right, then
     * we'll get a no senders notification and deallocate the receive right then, so all we destroy
     * here is the send right.
     */

    assert(local_port_type.at(ports[i]) == MACH_MSG_TYPE_PORT_RECEIVE);
    mach_call (mach_port_mod_refs (mach_task_self(), ports[i],
                                   MACH_PORT_RIGHT_SEND, -1));

    ports[i] = (~ remote_ports_by_local[ports[i]]);
  \}
else
  \{
    /* We've got a send right (note it), but no mapping to a remote port (the remote takes care of that) */

    local_port_type[ports[i]] = MACH_MSG_TYPE_PORT_SEND;

    /* request a DEAD NAME notification */

    mach_call (mach_port_request_notification (mach_task_self (), ports[i],
                                               MACH_NOTIFY_DEAD_NAME, 0,
                                               notification_port, MACH_MSG_TYPE_MAKE_SEND_ONCE, &old));
  \}
\end{semiverbatim}
\end{frame}

\begin{frame}[fragile]
\frametitle{{\tt netmsg}: Translation of a send right on reception}
\begin{semiverbatim}
\tiny
/* We're receiving a send right over the network.  We need to relay on a send
 * right, which will be to a local receive right.
 */

if (local_ports_by_remote.count(port) == 1)
  \{
    const mach_port_t newport = local_ports_by_remote[port];

    assert(local_port_type.at(newport) == MACH_MSG_TYPE_PORT_RECEIVE);

    /* it already exists; create a new send right to relay on */
    mach_call (mach_port_insert_right (mach_task_self (), newport, newport,
                                       MACH_MSG_TYPE_MAKE_SEND));

    return newport;
  \}
else
  \{
    mach_port_t newport;

    /* create new receive port and a new send right that will be moved to the recipient */
    mach_call (mach_port_allocate (mach_task_self (), MACH_PORT_RIGHT_RECEIVE, &newport));
    mach_call (mach_port_insert_right (mach_task_self (), newport, newport,
                                       MACH_MSG_TYPE_MAKE_SEND));

    /* request notification when all send rights have been destroyed */
    mach_port_t old;
    mach_call (mach_port_request_notification (mach_task_self (), newport,
                                               MACH_NOTIFY_NO_SENDERS, 0,
                                               newport, MACH_MSG_TYPE_MAKE_SEND_ONCE, &old));
    return newport;
  \}
\end{semiverbatim}
\end{frame}

@@

\begin{comment}
\begin{frame}
\frametitle{What kind of bugs have I found?}
\begin{itemize}
\item failure to update pagemap entry
\item FLUSHING flag
\item DEAD NAME notifications
\item condition variable
\item weak refs
\end{itemize}
\end{frame}
\end{comment}

\begin{frame}
\frametitle{Difficulty tracking which clients have access}

% Because of the kernel's ability to proactively flush pages (unless PRECIOUS is specified),
% we can't be certain that a client on a page's ACCESSLIST actually possesses that page.
% All we know for sure is that a client not on ACCESSLIST does not possess the page.
